Systematic Misestimation of Machine Learning Performance in Neuroimaging Studies of Depression

### 2021-05-03 ms.tex
Updated to version of final NPP pupblication

\documentclass{article}

\usepackage{arxiv}
\usepackage{graphicx}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{siunitx}
\sisetup{group-separator = {,},group-minimum-digits=4}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac,xfrac}
\usepackage{microtype}
\usepackage[]{authblk}
\usepackage{url}
\usepackage[format=hang]{caption}
\usepackage{pgf}
\usepackage{multirow, bigdelim}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{placeins}
\usepackage{import}
\usepackage{subcaption}
%\usepackage{tikz}

%\definecolor{lime}{HTML}{A6CE39}
%\DeclareRobustCommand{\orcidicon}{%
%	\begin{tikzpicture}
%	\draw[lime, fill=lime] (0,0)
%	circle [radius=0.16]
%	node[white] {{\fontfamily{qag}\selectfont \tiny ID}};
%	\draw[white, fill=white] (-0.0825,0.095)
%	circle [radius=0.007];
%	\end{tikzpicture}
%	\hspace{-2mm}
%}
%
%
%\newcommand{\orcid}[1]{\href{https://orcid.org/#1}{\orcidicon}}

\renewcommand*{\Affilfont}{\footnotesize}
\usepackage{fancyhdr}

\pagestyle{fancy}
\chead{Performance Overestimation in Neuroimaging Studies of Depression}

\title{Systematic Overestimation of Machine Learning Performance in Neuroimaging Studies of Depression}
%\author[$\;\;$1,2]{\orcid{0000-0001-5164-8227}Claas Flint \thanks{indicates that the authors contributed equally to the work and should be regarded as first authors.}}
%\author[$*$ 3,5]{\orcid{0000-0002-3353-8566}Micah Cearns}
%\author[1]{\orcid{0000-0003-4749-3298}Nils Opel}
%\author[1]{\orcid{0000-0002-7018-4525}Ronny Redlich}
%\author[1]{\orcid{0000-0001-6587-2617}David M. A. Mehler}
%\author[1]{\orcid{0000-0001-7459-6634}Daniel Emden}
%\author[1]{\orcid{0000-0002-6241-1492}Nils R. Winter}
%\author[1]{Ramona Leenings}
%\author[4,8]{\orcid{0000-0001-6363-2759}Simon B. Eickhoff}
%\author[6]{\orcid{0000-0002-2514-2625}Tilo Kircher}
%\author[6]{\orcid{0000-0002-0564-2497}Axel Krug}
%\author[6]{\orcid{0000-0002-0749-7473}Igor Nenadic}
%\author[1]{\orcid{0000-0002-2445-9778}Volker Arolt}
%\author[3]{Scott Clark}
%\author[3,5,7]{\orcid{0000-0001-6548-426X}Bernhard T. Baune}
%\author[2]{\orcid{0000-0001-7678-9528}Xiaoyi Jiang}
%\author[$\;\;$1]{\orcid{0000-0002-0623-3759}Udo Dannlowski \thanks{indicates that the authors contributed equally to the work and should be regarded as senior authors.}}
%\author[$\;\dagger \ddagger$1]{\orcid{0000-0002-8929-4134}Tim Hahn}

\author[$\;\;$1,2]{Claas Flint \thanks{indicates that the authors contributed equally to the work and should be regarded as first authors.}}
\author[$*$ 3,5]{Micah Cearns}
\author[1]{Nils Opel}
\author[1]{Ronny Redlich}
\author[1]{David M. A. Mehler}
\author[1]{Daniel Emden}
\author[1]{Nils R. Winter}
\author[1]{Ramona Leenings}
\author[4,8]{Simon B. Eickhoff}
\author[6]{Tilo Kircher}
\author[6]{Axel Krug}
\author[6]{Igor Nenadic}
\author[1]{Volker Arolt}
\author[3]{Scott Clark}
\author[3,5,7]{Bernhard T. Baune}
\author[2]{Xiaoyi Jiang}
\author[$\;\;$1]{Udo Dannlowski \thanks{indicates that the authors contributed equally to the work and should be regarded as senior authors.}}
\author[$\;\dagger \ddagger$1]{Tim Hahn}

\affil[1]{Department of Psychiatry, University of Münster, Germany}
\affil[2]{Faculty of Mathematics and Computer Science, University of Münster, Germany}
\affil[3]{Discipline of Psychiatry, School of Medicine, University of Adelaide, Australia}
\affil[4]{Institute of Neuroscience and Medicine (INM-7) Research Center Jülich}
\affil[5]{Department of Psychiatry, Melbourne Medical School, The University of Melbourne, Parkville, Australia}
\affil[6]{Department of Psychiatry and Psychotherapy, University of Marburg, Germany}
\affil[7]{The Florey Institute of Neuroscience and Mental Health, The University of Melbourne, Parkville, Australia}
\affil[8]{Institute of Systems Neuroscience, Medical Faculty, Heinrich Heine University Düsseldorf, Düsseldorf, Germany}
\affil[$\ddagger$]{Corresponding author: Tim Hahn, Phone: +49-251-83-56610, Fax: +49-251-8356612, Email: \href{mailto:hahnt@uni-muenster.de}{hahnt@uni-muenster.de}}

\begin{document}
    \maketitle
    \begin{abstract}
        We currently observe a disconcerting phenomenon in machine learning studies in psychiatry: While we would expect larger samples to yield better results due to the availability of more data, larger machine learning studies consistently show much weaker performance than the numerous small-scale studies. Here, we systematically investigated this effect focusing on one of the most heavily studied question in the field, namely the classification of patients suffering from Major Depressive Disorder (MDD) and healthy controls. Drawing upon a balanced sample of $N = \num{1868}$ MDD patients and healthy controls from our recent international Predictive Analytics Competition (PAC), we first trained and tested a classification model on the full dataset which yielded an accuracy of \SI{61}{\percent}. Next, we mimicked the process by which researchers would draw samples of various sizes ($N=4$ to $N=150$) from the population and showed a strong risk of overestimation. Specifically, for small sample sizes ($N=20$), we observe accuracies of up to \SI{95}{\percent}. For medium sample sizes ($N=100$) accuracies up to \SI{75}{\percent} were found. Importantly, further investigation showed that sufficiently large test sets effectively protect against performance overestimation whereas larger datasets per se do not. While these results questions the validity of a substantial part of the current literature, we outline the relatively low-cost remedy of larger test sets.
    \end{abstract}

    % keywords can be removed
    \keywords{machine learning \and neuroimaging \and major depression \and overestimation \and small sample size \and clinical translation}

    \pagebreak
    \section{Introduction}
    In psychiatry, we are witnessing an explosion of interest in machine learning (ML) and artificial intelligence for prediction and biomarker discovery, paralleling similar developments in personalized medicine\cite{Darcy2016,Eyre2016,Gabrieli2015,Jordan2015}. In contrast to the majority of investigations employing classic group-level statistical inference, ML approaches aim to build models which allow for individual (i.e. single subject) predictions, thus enabling direct assessment of individual differences and clinical utility\cite{Hahn2017}. While this constitutes a major advancement for clinical translation, recent results of large-scale investigations have given rise to a fundamental concern in the field: Specifically, machine learning studies including larger samples did not yield stronger performance, but consistently showed weaker results than studies drawing on small samples, calling into question the validity and generalisability of a large number of highly published proof-of-concept studies.

    The magnitude of this issue was impressively illustrated by the results of the Predictive Analytics Competition (PAC 2018; \url{http://photon-ai.com/pac}) in which participants developed machine learning models to classify healthy controls (HC) and depressive patients (MDD) based on structural MRI data from $N=\num{2240}$ participants. Despite the best efforts of $~\num{170}$ machine learners in \num{49} teams from around the world, performance ranged between \SI{60}{\percent} and \SI{65}{\percent} percent accuracy in a large, independent test set. This is in strong contrast to the numerous smaller studies showing accuracies of \SI{80}{\percent} or more\cite{Johnston2015,Mwangi2012,Patel2015}.
    Further empirical studies focusing on other disorders support this observed effect of performance deterioration with increasing sample size: In a large-scale investigation, Neuhaus \& Popescu\cite{Neuhaus2018} aggregated original studies across disorder categories, including schizophrenia (total observation $N=\num{5563}$), MDD ($N=\num{2042}$), and attention deficit hyperactivity disorder (ADHD, $N=\num{8084}$), finding an inverse relationship between sample size and balanced accuracy (schizophrenia, $r=\num{-.34}$; MDD, $r=\num{-.32}$; and ADHD, $r=\num{-.43}$). Similar results were observed in a recent review of \num{200} neuroimaging classification studies of brain disorders, finding a general trend towards lower reported accuracy scores in larger samples\cite{Arbabshirani2017}. Given that model performance would be expected to increase with more data, these results hint at a fundamental issue hampering current predictive biomarker studies in psychiatry.

    From a methodological point of view, it has been known since the early 90’s that training samples should be large when there is a high number of features (i.e. measured variables) and a complex classification rule being fit to a dataset\cite{Raudys1991}. Recent works have further reiterated this point\cite{VanderPloeg2014}. Moreover, it is possible that these effects have been further amplified by certain cross-validation schemes. For example, Kambeitz et al.\cite{Kambeitz2017} observed higher accuracy estimates in studies using hold-out cross-validation strategies compared to 10-fold and leave-one-out (LOOCV), whilst Varoquaux et al observed that LOOCV leads to unstable and biased estimates, concluding that repeated random splits should be preferred\cite{Varoquaux2017}.

    Although these findings have sparked in-depth conceptual considerations\cite{Hahn2019}, empirical investigations of this problem have been limited to specific cross-validation schemes and small test set sizes\cite{Varoquaux2018}. Here, we aim to systematically investigate the effects of both train and test set sample sizes on machine learning model performance in neuroimaging based MDD classification. In addition, as it is possible that effects of systematic overestimation have arisen due to suboptimal pipeline configurations (i.e., the disproportionate use of LOOCV and linear Support Vector Machines on samples containing more predictors than observations), we also test a further \num{48} different pipeline configurations to quantify the influence of these additional factors. To demonstrate that this effect was not dependent on the data or any pipeline configurations used in our analyses, we repeat the process using a dummy classifier that ignores all input variables and attempts to only predict class (outcome) proportions. To quantify the magnitude of these effects in each configuration, we drew samples of various sizes from the PAC dataset – mimicking the process by which researchers would draw samples from the population of ML studies reported in the literature. The resulting probability distributions are investigated.

    \section{Materials and methods}
    To investigate the effects of both train and test set sample sizes on machine learning model performance in neuroimaging based MDD classification, we repeatedly drew samples of different sizes from the PAC dataset to imitate the procedure reported in the literature. Subsequently, the resulting probability distributions are investigated.

    \subsection{Data description}
    The PAC dataset comprised anonymized, pre-processed MRI data of $N=\num{2240}$ individuals obtained from two large, independent, ongoing studies - the Münster Neuroimaging Cohort\cite{Dannlowski2015,Dannlowski2015a} ($N=\num{724}$ MDD; $N=\num{285}$ HC) and the FOR2107-study\cite{Kircher2018} ($N=\num{582}$ MDD, $N=\num{649}$ HC). Case/control status was diagnosed with the SCID-IV\cite{Wittchen1997} interview employed by trained clinical raters in both studies. In both cohorts, exclusion criteria were any neurological or severe medical condition, MRI contraindications, substance related disorders, Benzodiazepine treatment and, head injuries. For healthy controls, any current or previous psychiatric disorder or use of any psychotropic substances. The Münster Neuroimaging Cohort was scanned at one single MRI-scanner with the same sequence, while the FOR2107-study was scanned at two independent sites\cite{Vogelbacher2018}, yielding \num{3} different scanner types/sequences. The structural T1-weighted magnetic resonance imaging (MRI) scans were pre-processed with the CAT12 toolbox (\url{http://www.neuro.uni-jena.de/cat}, r1184) using default parameters to obtain modulated, normalized grey matter segments (resolution $\num{1.5} \times \num{1.5} \times \SI{1.5}{\cubic\milli\meter}$) which were used for the present analysis. Furthermore, age, gender, scanner type, and total intracranial volume (TIV) were provided.

    \subsection{Machine learning pipeline}
    To ensure the unbiased approximation of the model’s performance in previously unseen patients (i.e. model generalization), we trained and tested all models in a pipeline to prevent information leaking between patients used for training and validation. First, we used random under sampling to derive a balanced sample of \num{934} MDD cases and \num{934} healthy controls (HC) (see Supplementary appendix \ref{cha:summary_statistics} for summary statistics). To reduce the computational effort, all images have been scaled down to a voxel size of  $\num{3} \times \num{3} \times \SI{3}{\cubic\milli\meter}$. Following, every image was converted to a vector, where every voxel served as a feature. For standardization, the features were scaled to have zero mean and unit variance. Finally, a linear support vector machine (SVM) with default parameters (sklearn\cite{Scikit-learn2018}, v0.20) was trained and model performance was calculated and analysed in the subsequent analyses. The effect of varying scanner-distributions across samples was found to be negligible (see Supplementary appendix \ref{cha:influence_of_scanner_distribution}).

    \subsection{Overall sample size effects}
    To first examine the effects of overall-set size, we randomly sampled the PAC dataset in steps of \num{1} from $N=4$ to $N=150$. For each $N$ value we created \num{1000} balanced samples. This yielded \num{147000} random samples with equal numbers of patients and healthy controls. Applying the SVM pipeline described above to each of these samples independently allowed us to obtain a distribution of accuracy scores for each $N$ value, respectively. On each sample one SVM was trained with default parameters using LOOCV. Thus, we trained a total of \num{11315000} SVMs. Since the computational effort increases quadratically with increasing sample size, and in addition, the most recent neuroimaging ML studies using LOOCV rarely exceed $N=150$, we decided to stop at this value. Following, we evaluated the distribution of accuracy scores estimated for each sample size ($N=4$ to $150$).

    \subsection{Training set size effects}
    Second, to examine the effects of training set size, we varied the size of each training set and then tested performance on a fixed hold-out set. Specifically, we randomly sampled the PAC dataset in steps of $1$ from $N=4$ to $150$ as was done in the previous analysis. For each $N$ value, we created \num{1000} balanced samples and used each of them to train different models. To then test the effects of training set size on test set performance, we tested each trained model on a balanced test set of $N=300$. From the resulting distributions, we quantified the probability of overestimating accuracy as a function of training set size.

    \subsection{Test set size effects}
    To assess the effects of test set size on classification accuracy, we randomly sampled a balanced group of \num{300} subjects from our full sample of \num{934} MDD cases and \num{934} HC’s. From this sample, we randomly sub-sampled test sets of $N=4$ to $150$ in steps of $1$. For each value of $N$, we took \num{1000} samples, resulting in the creation of \num{147000} random test samples. From the remaining \num{784} MDD cases and \num{784} HC’s, we took out another \SI{20}{\percent} sample ($\text{MDD}=157$ and $\text{HC}=157$). The prediction of this sample provides a reliable basis for an overall performance estimation and allows for a comparison with the results obtained from smaller samples. We then trained a single SVM on the remaining sample ($\text{MDD}=627$, $\text{HC}=627$). The trained SVM was then tested on each test set sample ($N=4$ to $150$). From the resulting test set accuracy distributions, we derived the probability of obtaining accuracy scores between \SI{50}{\percent} and \SI{90}{\percent} accuracy by chance. See Figure \ref{fig:pipeline} for an overview of all analyses.

    \begin{figure}
        \def\svgwidth{\textwidth}
        \import{./images/}{coloured_horizontal_sys_over_pipeline.pdf_tex}
        \caption{Workflow for overall, train, and test set size systematic overestimation analyses.}
        \label{fig:pipeline}
    \end{figure}

    \subsection{Generalizability of statistical effect}
    \subsubsection{Alternative pipeline configurations}
    As we have attempted to hold other components of the modelling process constant so any observed effects of systematic overestimation can be attributed to sample size alone, it is also possible that our results may be dependent on the basic configuration of our machine learning pipeline (e.g., the use of a linear SVM, default hyperparameters, and LOOCV). Therefore, we trained a further \num{48} pipeline configurations, including the use of both linear and radial basis function SVMs and a Random Forest classifier, all of which have demonstrated their efficacy in neuroimaging classification studies.\cite{Kambeitz2017} Within these configurations, we conducted dimensionality reduction using principal components analysis as well as f-test based feature selection. This allowed us to assess whether our findings were being confounded by the large number of predictors used in the main analysis. For further information see Supplement Appendix \ref{cha:alternative_machine_configurations} and for all results see supplementary results \ref{cha:alternative_machine_configuration_results}, supplementary figures \ref{fig:no_PCA_no_selection_RandomForest} – \ref{fig:PCA_all_components_1000_best_selected_LinearSVC} and supplementary tables \ref{tab:no_PCA_no_selection_RandomForest} – \ref{tab:PCA_all_components_1000_best_selected_LinearSVC}.
    Dummy classifier
    In order to show that the observed effects were not specific to the PAC dataset or any of the alternative pipeline configurations in these analyses, we repeated the procedures described above with a dummy classifier. Our dummy classifier assumed a prior probability for MDD vs control classification based on the percentage proportion of each class in the training data (prevalence). As the dataset was balanced with random under sampling, the prior and subsequent ground truth of the model was equal to exactly \SI{50}{\percent}. This approach allowed us to compare our distribution of dummy performance estimates derived from our subsample analysis to this ground truth value. Importantly, this approach allows for the quantification of accuracy overestimation as a function of sample size completely independent of any unique characteristics that may be specific to the PAC dataset or our pipeline configurations. Therefore, we can then be sure that any subsequent changes in classifier performance are attributable to sample size alone.

>>>>

\documentclass[12pt,a4paper]{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{subcaption}
\usepackage{siunitx}
\sisetup{group-separator = {,},group-minimum-digits=4}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{bm}
\newcommand*{\B}[1]{\ifmmode\bm{#1}\else\textbf{#1}\fi}
\usepackage{nicefrac,xfrac}
\usepackage[]{authblk}
\usepackage[format=hang]{caption}
\usepackage{multirow, bigdelim}
\usepackage{xcolor}
\usepackage[linkbordercolor={0.818 0.394 0.394},urlbordercolor={0.366 0.528 0.745},hidelinks]{hyperref}
\usepackage{placeins}
\usepackage{import}
\usepackage[printonlyused, nohyperlinks]{acronym}
\usepackage{csquotes}
\usepackage[super]{nth}
\usepackage{blindtext}
\usepackage{float}

\renewcommand*{\Affilfont}{\footnotesize}

\title{Systematic Misestimation of Machine Learning Performance in Neuroimaging Studies of Depression}

\author[$\;\;$1,2]{\href{https://orcid.org/0000-0001-5164-8227}{Claas Flint \thanks{indicates that the authors contributed equally to the work and should be regarded as first authors.}}}
\author[$*$ 3,5]{\href{https://orcid.org/0000-0002-3353-8566}{Micah Cearns}}
\author[1]{\href{https://orcid.org/0000-0003-4749-3298}{Nils Opel}}
\author[1]{\href{https://orcid.org/0000-0002-7018-4525}{Ronny Redlich}}
\author[1]{\href{https://orcid.org/0000-0001-6587-2617}{David M. A. Mehler}}
\author[1]{\href{https://orcid.org/0000-0001-7459-6634}{Daniel Emden}}
\author[1]{\href{https://orcid.org/0000-0002-6241-1492}{Nils R. Winter}}
\author[1]{Ramona Leenings}
\author[4,8]{\href{https://orcid.org/0000-0001-6363-2759}{Simon B. Eickhoff}}
\author[6]{\href{https://orcid.org/0000-0002-2514-2625}{Tilo Kircher}}
\author[6]{\href{https://orcid.org/0000-0002-0564-2497}{Axel Krug}}
\author[6]{\href{https://orcid.org/0000-0002-0749-7473}{Igor Nenadic}}
\author[1]{\href{https://orcid.org/0000-0002-2445-9778}{Volker Arolt}}
\author[3]{Scott Clark}
\author[3,5,7]{\href{https://orcid.org/0000-0001-6548-426X}{Bernhard T. Baune}}
\author[2]{\href{https://orcid.org/0000-0001-7678-9528}{Xiaoyi Jiang}}
\author[$\;\;\ddagger$1]{\href{https://orcid.org/0000-0002-0623-3759}{Udo Dannlowski \thanks{indicates that the authors contributed equally to the work and should be regarded as senior authors.}}}
\author[$\;\dagger$1]{\href{https://orcid.org/0000-0002-8929-4134}{Tim Hahn}}

\affil[1]{Department of Psychiatry, University of Münster, Germany}
\affil[2]{Faculty of Mathematics and Computer Science, University of Münster, Germany}
\affil[3]{Discipline of Psychiatry, School of Medicine, University of Adelaide, Australia}
\affil[4]{Institute of Neuroscience and Medicine (INM-7) Research Center Jülich}
\affil[5]{Department of Psychiatry, Melbourne Medical School, The University of Melbourne, Parkville, Australia}
\affil[6]{Department of Psychiatry and Psychotherapy, University of Marburg, Germany}
\affil[7]{The Florey Institute of Neuroscience and Mental Health, The University of Melbourne, Parkville, Australia}
\affil[8]{Institute of Systems Neuroscience, Medical Faculty, Heinrich Heine University Düsseldorf, Düsseldorf, Germany}
\affil[$\ddagger$]{Corresponding author: Udo Dannlowski, Phone: +49-251-83-56610, Email: \href{mailto:dannlow@uni-muenster.de}{dannlow@uni-muenster.de}}

\renewcommand{\shorttitle}{Performance Misestimation in Neuroimaging Studies of Depression}
\hypersetup{
    pdftitle={Systematic Misestimation of Machine Learning Performance in Neuroimaging Studies of Depression},
    pdfauthor={
        Claas Flint,
        Micah Cearns,
        Nils Opel,
        Ronny Redlich,
        David M. A. Mehler,
        Daniel Emden,
        Nils R. Winter,
        Ramona Leenings,
        Simon B. Eickhoff,
        Tilo Kircher,
        Axel Krug,
        Igor Nenadic,
        Volker Arolt,
        Scott Clark,
        Bernhard T. Baune,
        Xiaoyi Jiang,
        Udo Dannlowski,
        Tim Hahn
    },
    pdfproducer={Claas Flint},
    pdfkeywords={machine learning, neuroimaging, major depression, misestimation, small sample size, clinical translation},
    colorlinks=false
}

%! language = Latex
\begin{document}
    \maketitle
    \begin{abstract}
        We currently observe a disconcerting phenomenon in machine learning studies in psychiatry: While we would expect larger samples to yield better results due to the availability of more data, larger machine learning studies consistently show much weaker performance than the numerous small-scale studies. Here, we systematically investigated this effect focusing on one of the mostheavily studied questions in the field, namely the classification of patients suffering from \ac{mdd} and \ac{hc} based on neuroimaging data. Drawing upon structural \ac{mri} data from a balanced sample of $N$\,=\;\num{1868} \ac{mdd} patients and \ac{hc} from our recent international \ac{pac}, we first trained and tested a classification model on the full dataset which yielded an accuracy of \SI{61}{\percent}. Next, we mimicked the process by which researchers would drawsamples of various sizes ($N$\,=\;\num{4} to $N$\,=\;\num{150}) from the population and showed a strong risk of misestimation. Specifically, for smallsample sizes ($N$\,=\;\num{20}), we observe accuracies of up to \SI{95}{\percent}. For medium sample sizes ($N$\,=\;\num{100}) accuracies up to \SI{75}{\percent} were found. Importantly, further investigation showed that sufficiently large test sets effectively protect against performance misestimationwhereas larger datasets per se do not. While these results question the validity of a substantial part of the current literature, weoutline the relatively low-cost remedy of larger test sets, which is readily available in most cases.

    \end{abstract}

    \keywords{machine learning \and neuroimaging \and major depression \and overestimation \and small sample size \and clinical translation}

    \pagebreak

    \section{Introduction}
    \acresetall
    In psychiatry, we are witnessing an explosion of interest in \ac{ml} and artificial intelligence for prediction and biomarker discovery, paralleling similar developments in personalized medicine~\cite{Darcy2016,Eyre2016,Gabrieli2015,Jordan2015}. In contrast to the majority of investigations employing classic group-level statistical inference, \ac{ml} approaches aim to build models which allow for individual (i.e.\ single subject) predictions, thus enabling direct assessment of individual differences and clinical utility~\cite{Hahn2017}. While this constitutes a major advancement for clinical translation, recent results of large-scale investigations have given rise to a fundamental concern in the field: Specifically, \ac{ml} studies including larger samples did not yield stronger performance, but consistently showed weaker results than studies drawing on small samples, calling into question the validity and generalizability of a large number of highly published proof-of-concept studies.

    The magnitude of this issue was impressively illustrated by the results of the \ac{pac} 2018 (Supplementary appendix~/ref{cha:pac-2018}) in which participants developed \ac{ml} models to classify \ac{hc} and patients diagnosed with \ac{mdd} based on structural \ac{mri} data from $N=\num{2240}$ participants. Despite the best efforts of $\sim\num{170}$ machine learners in \num{49} teams from around the world, performance ranged between \SI{60}{\percent} and \SI{65}{\percent} accuracy in a large, independent test set. This is in strong contrast to the numerous smaller studies showing accuracies of \SI{80}{\percent} or more~\cite{Johnston2015,Mwangi2012,Patel2015}.

    Further empirical studies focusing on other disorders support this observed effect of performance deterioration with increasing sample size: In a large-scale investigation, Neuhaus \& Popescu~\cite{Neuhaus2018} aggregated original studies across disorder categories, including schizophrenia(total observation $N=\num{5563}$), \ac{mdd} ($N=\num{2042}$), and \ac{adhd} ($N=\num{8084}$), finding an inverse relationship between sample size and balanced accuracy(schizophrenia, $r=\num{-.34}$; \ac{mdd}, $r=\num{-.32}$; and \ac{adhd}, $r=\num{-.43}$). Similar results were observed in a recent review of \num{200} neuroimaging classification studies of brain disorders, finding a general trend towards lower reported accuracy scores in larger samples~\cite{Arbabshirani2017}. Given that model performance would be expected to increase with more data, these results hint at a fundamental issue hampering current predictive biomarker studies in psychiatry.

    From a methodological point of view, it has been known since the early 90's that training samples should be large when there is a high number of features (i.e.\ measured variables) and a complex classification rule being fit to a dataset~\cite{Raudys1991}. Recent works have further reiterated this point~\cite{VanderPloeg2014}. Moreover, these effects may have been further amplified by certain cross-validation schemes. For example, Kambeitz et al.~\cite{Kambeitz2017} observed higher accuracy estimates in studies using hold-out cross-validation strategies compared to 10-fold and \ac{loocv}, whilst Varoquaux et al. observed that \ac{loocv} leads to unstable and biased estimates, concluding that repeated random splits should be preferred~\cite{Varoquaux2017}.

    Although these findings have sparked in-depth conceptual considerations~\cite{Hahn2019}, empirical investigations of this problem have been limited to specific cross-validation schemes and small test set sizes~\cite{Varoquaux2018}. Here, we aim to systematically investigate the effects of both train and test set sample sizes on \ac{ml} model performance in neuroimaging based \ac{mdd} classification. In addition, as it is possible that effects of systematic misestimation have arisen due to suboptimal pipeline configurations (i.e., the disproportionate use of \ac{loocv} and linear \acp{svm} on samples containing more predictors than observations), we also test a further \num{48} different pipeline configurations to quantify the influence of these additional factors. To demonstrate that this effect was not dependent on the data or any pipeline configurations used in our analyses, we repeat the process using a \emph{dummy classifier}. To quantify the magnitude of these effects in each configuration, we drew samples of various sizes from the \ac{pac} dataset - mimicking the process by which researchers would draw samples from the population of \ac{ml} studies reported in the literature. The resulting probability distributions are investigated.

    \section{Materials and methods}
    To investigate the effects of both train and test set sample sizes on \ac{ml} model performance in neuroimaging based \ac{mdd} classification, we repeatedly drew samples of different sizes from the \ac{pac} dataset to imitate the procedure reported in the literature. Subsequently, the resulting probability distributions are investigated.

    \subsection{Data description}
    The \ac{pac} dataset comprised anonymized, pre-processed \ac{mri} data of $N=\num{2240}$ individuals obtained from two large, independent, ongoing studies - the Münster Neuroimaging Cohort~\cite{Dannlowski2015,Dannlowski2015a} ($N=\num{724}$ \ac{mdd}; $N=\num{285}$ \ac{hc}) and the FOR2107-study~\cite{Kircher2018} ($N=\num{582}$ \ac{mdd}, $N=\num{649}$ \ac{hc}). Case/control status was diagnosed with the {SCID-IV}~\cite{Wittchen1997} interview employed by trained clinical raters in both studies. In both cohorts, exclusion criteria were any neurological or severe medical condition, \ac{mri} contraindications, substance-related disorders, Benzodiazepine treatment and head injuries. For healthy controls, any current or previous psychiatric disorder or use of any psychotropic substances. The Münster Neuroimaging Cohort was scanned at one single \ac{mri}-scanner with the same sequence, while the FOR2107-study was scanned at two independent sites~\cite{Vogelbacher2018}, yielding \num{3} different scanner types/sequences. The structural T1-weighted \ac{mri} scans were pre-processed with the CAT12 toolbox (\url{http://www.neuro.uni-jena.de/cat}, r1184) using default parameters to obtain modulated, normalized grey matter segments (resolution $\num{1.5} \times \num{1.5} \times \SI{1.5}{\cubic\milli\meter}$) which were used for the present analysis. Furthermore, age, gender, scanner type, and \ac{tiv} were provided.

    The FOR2107 cohort project was approved by the Ethics Committees of the Medical Faculties, University of Marburg and University of Münster.

    \subsection{Machine learning pipeline}
    To ensure the unbiased approximation of the model's performance in previously unseen patients (i.e., model generalization), we trained and tested all models in a pipeline to prevent information leaking between patients used for training and validation. To avoid a confounding effect due to an imbalance in the sample, we used random under-sampling in a first step to obtain a balanced sample of \num{934} \ac{mdd} cases and \num{934} \ac{hc} (see Supplementary appendix~\ref{cha:summary-statistics} for summary statistics). To reduce the computational effort, all images have been scaled down to a voxel size of $\num{3} \times \num{3} \times \SI{3}{\cubic\milli\meter}$. Following, every image was converted to a vector, where every voxel served as a feature. After background elimination (features with no variance), \num{58255} features remained. For standardization, the features were scaled to have zero mean and unit variance. Finally, a linear \ac{svm} with default parameters (Scikit-learn~\cite{scikit-learn}, v0.20) was trained and model performance was calculated and analysed in the subsequent analyses. The effect of varying scanner-distributions across samples was found to be negligible (see Supplementary appendix~\ref{cha:influence-of-scanner-distribution}).

    \subsection{Overall sample size effects}
    To first examine the effects of overall-set size, we randomly sampled the \ac{pac} dataset in steps of \num{1} from $N=4$ to $N=150$. For each $N$ value we created \num{1000} balanced samples. This yielded \num{147000} random samples with equal numbers of \ac{mdd} patients and \ac{hc}. Applying the \ac{svm} pipeline described above to each of these samples independently allowed us to obtain a distribution of accuracy scores for each $N$ value, respectively. On each sample one \ac{svm} was trained with default parameters using \ac{loocv}. Thus, we trained a total of \num{11315000} \acp{svm}. Since the computational effort increases quadratically with increasing sample size, and in addition, the most recent neuroimaging \ac{ml} studies using \ac{loocv} rarely exceed $N=150$, we decided to stop at this value. Following, we evaluated the distribution of accuracy scores estimated for each sample size ($N=4$ to $150$).

    \subsection{Training set size effects}
    Second, to examine the effects of training set size, we varied the size of each training set and then tested performance on a fixed hold-out set. Specifically, we randomly sampled the \ac{pac} dataset in steps of $1$ from $N=4$ to $150$ as was done in the previous analysis. For each $N$ value, we created \num{1000} balanced samples and used each of them to train different models. To then test the effects of training set size on test set performance, we tested each trained model on a balanced test set of $N=300$. From the resulting distributions, we quantified the probability of overestimating accuracy as a function of training set size.

    \subsection{Test set size effects}
    To assess the effects of test set size on classification accuracy, we randomly sampled a balanced group of \num{300} subjects from our full sample of \num{934} \ac{mdd} cases and \num{934} \acp{hc}. From this sample, we randomly sub-sampled test sets of $N=\num{4}$ to \num{150} in steps of \num{1}. For each value of $N$, we took \num{1000} samples, resulting in the creation of \num{147000} random test samples. From the remaining \num{784} \ac{mdd} cases and \num{784} \acp{hc}, we took out another \SI{20}{\percent} sample ($\text{\ac{mdd}}=157$ and $\text{\ac{hc}}=157$). The prediction of this sample provides a reliable basis for an overall performance estimation and allows for a comparison with the results obtained from smaller samples. We then trained a single \ac{svm} on the remaining sample ($\text{\ac{mdd}}=627$, $\text{\ac{hc}}=627$). The trained \ac{svm} was then tested on each test set sample ($N=\num{4}$ to \num{150}). From the resulting test set accuracy distributions, we derived the probability of obtaining accuracy scores between \SI{50}{\percent} and \SI{90}{\percent} accuracy by chance. See Figure~\ref{fig:workflow} for an overview of all analyses.

    \begin{figure}
        \includegraphics[width=\textwidth]{./images/figure_1}
        \caption{Workflow to investigate the correlations between sample size and misestimation. First the effect of misestimation is investigated over the whole classification process ((1) Overall sample size analysis). Following the process of training and testing is evaluated seperatly ((2) Training set sample size analysis, (3) Test set sample size analysis)}
        \label{fig:workflow}
    \end{figure}

    \subsection{Generalizability of statistical effect}

    \subsubsection{Alternative pipeline configurations}
    As we have attempted to hold other components of the modelling process constant so any observed effects of systematic misestimation can be attributed to sample size alone, it is also possible that our results may be dependent on the basic configuration of our \ac{ml} pipeline (e.g., the use of a linear \ac{svm}, default hyperparameters, and \ac{loocv}). Therefore, we trained a further \num{48} pipeline configurations, including the use of both linear and radial basis function \acp{svm} and a Random Forest classifier, all of which have demonstrated their efficacy in neuroimaging classification studies~\cite{Kambeitz2017}. Within these configurations, we conducted dimensionality reduction using \ac{pca} as well as f-test based feature selection. This allowed us to assess whether our findings were being confounded by the large number of predictors used in the main analysis. For further information see Supplement Append~x~\ref{cha:alternative-machine-configurations} and for all results see supplementary results~\ref{cha:alternative-machine-configuration-results}, supplementary figures~\ref{fig:no_PCA_no_selection_RandomForest} -~\ref{fig:PCA_all_components_1000_best_selected_LinearSVC} and supplementary tables~\ref{tab:no_PCA_no_selection_RandomForest}~-~\ref{tab:PCA_all_components_1000_best_selected_LinearSVC}.

    \subsubsection{Dummy classifier}
In order to show that the observed effects were not specific to the \ac{pac} dataset or any of the alternative pipeline configurations in these analyses, we repeated the procedures described above with a dummy classifier. Our dummy classifier assumed a prior probability for \ac{mdd} vs \ac{hc} classification based on the percentage proportion of each class in the training data (prevalence). As the dataset was balanced with random under-sampling, the prior and subsequent ground truth of the model was equal to exactly \SI{50}{\percent}. This approach allowed us to compare our distribution of dummy performance estimates derived from our subsample analysis to this ground truth value. Importantly, this approach allows for the quantification of accuracy misestimation as a function of sample size completely independent of any unique characteristics that may be specific to the \ac{pac} dataset or our pipeline configurations. Therefore, we can then be sure that any subsequent changes in classifier performance are attributable to sample size alone.

    \section{Results}
